Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-22 WSSSPE1_report/WSSSPE1_summary_report.tex
Move CfP to Appendix and revise Process.

WSSSPE1 was organized in collaboration of the relatively small group
of five organizers and a larger peer-review committee with~36
members. This committee had early influence on the organization, e.g.,
on the specific call for papers.

The workshop call for papers included:

\begin{quote}
``In practice, scientific software activities are part of an ecosystem
where key roles are held by developers, users, and funders. All three
groups supply resources to the ecosystem, as well as requirements that
bound it. Roughly following the example of NSF's Vision and Strategy
for Software~\cite{NSF_software_vision},
the ecosystem may be viewed as having challenges related to:

\begin{itemize}[leftmargin=0.2in]
\item the development process that leads to new (versions of) software
\begin{itemize}[leftmargin=0.2in]
\item how fundamental research in computer science or
  science/engineering domains is turned into reusable software
\item software created as a by-product of research
\item impact of computer science research on the development of
    scientific software and vice versa
\end{itemize}
\item the support and maintenance of existing software, including
  software engineering
\begin{itemize}[leftmargin=0.2in]
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and
  sustainability
\end{itemize}
\item the role of open source communities or industry
\item use of the software
\begin{itemize}[leftmargin=0.2in]
\item growing communities
\item reproducibility, transparency needs that may be unique to science
\end{itemize}
\item policy issues, such as
\begin{itemize}[leftmargin=0.2in]
\item measuring usage and impact
\item software credit, attribution, incentive, and reward
\item career paths for developers and institutional roles
\item issues related to multiple organizations and multiple countries,
  such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of
  publishers
\end{itemize}
\item education and training''
\end{itemize}

\end{quote}

Based on the goal of encouraging a wide range of submissions from
those involved in software practice, ranging from initial thoughts and
partial studies to mature deployments, the organizers wanted to make
submission as easy as possible. The call for papers stated:

\begin{quote}
``We invite short (4-page) position/experience reports that will be used
to organize panel and discussion sessions. These papers will be
archived by a third-party service, and provided DOIs [Digital Object Identifiers].
We encourage
submitters to license their papers under a Creative Commons license
that encourages sharing and remixing, as we will combine ideas (with
attribution) into the outcomes of the workshop.  An interactive site
will be created to link these papers and the workshop discussion, with
options for later comments and contributions. Contributions will be
peer-reviewed for relevance and originality before the links are added
to the workshop site; contributions will also be used to determine
discussion topics and panelists. We will also plan one or more papers
to be collaboratively developed by the contributors, based on the
panels and discussions.''
\end{quote}

In total, 58 submissions were received, and almost all submitters used either
arXiv~\cite{arXiv-web} or Figshare~\cite{figshare-web} to self-publish
their papers.

A peer review process followed the submissions, where
the~58 papers received~181 reviews, an average of~3.12 reviews per
paper. Reviews were completed using a Google form, which allowed
reviewers to choose papers they wanted to review, then to provide
general comments and scores on relevance to the organizers and
to the authors. The review reports enabled the organizers to decide
which papers to associate with the workshop, and allowed the authors
to improve their papers.

The organizers decided to list~54 of the papers (see Appendix~\ref{sec:papers} for a list of these papers) as significant
contributions to the workshop, a very high acceptance rate but one
that is reasonable, given the goal of broad participation and the fact
that the reports were already self-published. Based on the submissions, the papers were
grouped into three main categories, namely \emph{Developing and
Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
category was associated with a panel and discussion at the
workshop.  Three to four representatives from each submission
category were appointed as panelists, and assigned to read a subset of
the papers in that category and then discuss them in the panel.
The workshop itself consisted of two keynote presentations followed by the
three panel/discussion sessions.  Notes were collaboratively taken
by the group during the discussion, using a web document~\cite{WSSSPE1-google-notes}.

@@

>>>>

WSSSPE1 was organized by a relatively small group
of five organizers and a larger program committee of~36
members. The program committee peer-reviewed submissions, but also had
early influence in the workshop's organization, such as articulating
the Call for Papers (see Appendix~\ref{sec:cfp}).

Aside from setting the stage for the relevance of software
sustainability and corresponding training and workforce issues to
science,  the call for papers enumerated the topics it was interested
in as challenges to the ecosystem of which scientific software is a
part, and in which software developers, users, and funders hold
roles. These challenges roughly followed NSF's Vision and Strategy
for Software~\cite{NSF_software_vision}, and specifically included the
development and research process that leads to new (or new versions of existing)
software; the support, community infrastructure, and engineering for
maintenance of existing software; the role of open source communities
and industry; aspects of the use of software, such as
reproducibility, that may be unique to science;  policy issues
related to software sustainability such as measuring impact,
giving credit, and incentivizing best practices; and education and training.

The workshop's goal was to encourage a wide range of submissions from
those involved in software practice, ranging from initial thoughts and
partial studies to mature deployments. Consequently, the organizers
aimed to make submission as easy as possible. Rather than requiring
adherence to a formal submission system and a full research
paper-style template, submissions were intentionally limited to short
4-page papers, articulating either a position on one or more of the
topics, or reporting experiences related to them. Furthermore, for
submission authors were asked to self-archive (and thus self-publish)
their papers with a third-party service that issues persistent
identifiers, such as Digital Object Identifiers (DOIs), and to then
submit the URL to the archived paper by email. This had the side
effect that every submitter would also have a publicly available and
citable version of their workshop contribution.

This process resulted in a total of 58 submissions. Almost all submitters used either
arXiv~\cite{arXiv-web} or Figshare~\cite{figshare-web} to archive
their papers. The submissions were then subjected to peer review by
the program committee, resulting in~181 reviews, an average of~3.12 reviews per
paper. Reviews were completed using a Google form, which allowed
reviewers to choose papers they wanted to review, and to provide
general comments as well as relevance scores to the organizers and
to the authors. Based on the review reports, the organizers decided to
list~54 of the papers (see Appendix~\ref{sec:papers} for a full list) as significant
contributions to the workshop. The high acceptance rate may come as a
surprise, but it is nonetheless consistent with the goal of
fostering broad participation, and as a corollary of the chosen submission
process paper acceptance was no longer a means to filter
the papers' public availability.

In roughly following the aforementioned scientific software ecosystem
challenges, the accepted submissions were
grouped into three main categories, namely \emph{Developing and
Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
category was assigned a panel, with three to four panelists 
drawn from authors of the associated submissions, who were each
assigned to read and summarily present a subset of
the papers associated with the panel. The process from organizing and
advertising the workshop, to collecting and reviewing the
papers, and putting together the agenda was documented by the
organizers in a report~\cite{WSSSPE1-pre-report}, which they
self-archived in the same way as contributed papers.

For each panel, the paper summaries were followed by active discussion
involving panelists, onsite attendees, and often enough online
participants. The latter was facilitated by having a shared Google
Doc~\cite{WSSSPE1-google-notes} for collaborative note taking. Some of
the online discussion also took place on Twitter (hashtag
\texttt{\#wssspe}).

The workshop itself consisted of two keynote presentations followed by the
three panel sessions.  Remote participation was
facilitated by a live-cast of all keynotes and panels on Ustream.tv (\url{http://ustream.tv}).

@@

\section{Call For Papers} \label{sec:cfp}

The full Call for Papers for WSSSPE1 can be found online at
\url{http://wssspe.researchcomputing.org.uk/wssspe1/cfp/}. We include
below the part that explains scope and topics for which submissions
were solicited (see \S\ref{sec:process} for number of submissions
received, accepted, and the corresponding process).

\begin{quote}
``Progress in scientific research is dependent on the quality and
accessibility of software at all levels and it is now critical to
address many new challenges related to the development, deployment,
and maintenance of reusable software. In addition, it is essential
that scientists, researchers, and students are able to learn and adopt
a new set of software-related skills and methodologies. Established
researchers are already acquiring some of these skills, and in
particular a specialized class of software developers is emerging in
academic environments who are an integral and embedded part of
successful research teams. This workshop will provide a forum for
discussion of the challenges, including both positions and
experiences. The short papers and discussion will be archived as a
basis for continued discussion, and we intend the workshop to feed
into the collaborative writing of one or more journal publications.

In practice, scientific software activities are part of an ecosystem
where key roles are held by developers, users, and funders. All three
groups supply resources to the ecosystem, as well as requirements that
bound it. Roughly following the example of NSF's Vision and Strategy
for Software~\cite{NSF_software_vision},
the ecosystem may be viewed as having challenges related to:

\begin{itemize}[leftmargin=0.2in]
\item the development process that leads to new (versions of) software
\begin{itemize}[leftmargin=0.2in]
\item how fundamental research in computer science or
  science/engineering domains is turned into reusable software
\item software created as a by-product of research
\item impact of computer science research on the development of
    scientific software and vice versa
\end{itemize}
\item the support and maintenance of existing software, including
  software engineering
\begin{itemize}[leftmargin=0.2in]
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and
  sustainability
\end{itemize}
\item the role of open source communities or industry
\item use of the software
\begin{itemize}[leftmargin=0.2in]
\item growing communities
\item reproducibility, transparency needs that may be unique to science
\end{itemize}
\item policy issues, such as
\begin{itemize}[leftmargin=0.2in]
\item measuring usage and impact
\item software credit, attribution, incentive, and reward
\item career paths for developers and institutional roles
\item issues related to multiple organizations and multiple countries,
  such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of
  publishers
\end{itemize}
\item education and training
\end{itemize}

This workshop is interested in all of the above topics.  We invite short (4-page) position/experience reports that will be used
to organize panel and discussion sessions. These papers will be
archived by a third-party service, and provided DOIs [Digital Object Identifiers].
We encourage
submitters to license their papers under a Creative Commons license
that encourages sharing and remixing, as we will combine ideas (with
attribution) into the outcomes of the workshop.  An interactive site
will be created to link these papers and the workshop discussion, with
options for later comments and contributions. Contributions will be
peer-reviewed for relevance and originality before the links are added
to the workshop site; contributions will also be used to determine
discussion topics and panelists. We will also plan one or more papers
to be collaboratively developed by the contributors, based on the
panels and discussions.''
\end{quote}
